---
title: "Final Project"
author: "Yanze Jiang"
date: "2025-05-12"
output: pdf_document
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(skimr)
library(janitor)
library(GGally)
library(caret)
library(rpart)
library(pROC)
library(ggplot2)
```

# Introduction

Customer churn occurs when a customer decides to stop using a company's services. In this 
report, the main focus is on bank customer churn. For banks, losing customers means losing 
long-term profits, so it is important to be able to predict customer churn so that banks can 
prepare for churn in advance. In this project, a dataset from Kaggle was used, which contains 
10,000 customer records with information such as credit score, account balance, whether the 
customer is active, and whether they have a credit card. The target variable is whether the 
customer churned (1 = yes, 0 = no). The hypothesis is that certain financial behaviors and 
account characteristics are associated with customer churn. Specifically, it is assumed that 
customers with lower credit scores, lower account balances, inactive accounts, or no credit cards 
are more likely to leave. The goal of this project is to test this hypothesis by building a machine 
learning model that can predict which customers are at risk of churn. 

# Data Cleaning

To begin, the dataset was imported and cleaned using packages like janitor and tidyverse. The 
customer_id column was removed since it does not contribute useful information for prediction. 
Categorical variables like gender and country were also dropped to focus on numeric features 
only. The target variable churn was converted into a factor with two levels: “no” and “yes”. 
The dataset was split into training (80%) and testing (20%) sets using createDataPartition() from 
the caret package. Median imputation was applied to handle any missing values in the training 
set. Finally, the structure of the data was checked to make sure all variables were in the correct 
format for modeling. 

```{r, echo=FALSE, warning=FALSE, message=FALSE}
plot_roc_auc <- function(model, test_data, response_var) {
  probs <- predict(model, newdata = test_data, type = "prob")[, 2]
  
  true_labels <- test_data[[response_var]]
  
  if (is.factor(true_labels)) {
    true_labels <- as.numeric(true_labels) - 1
  }
  
  roc_obj <- roc(response = true_labels, predictor = probs)
  
  auc_val <- auc(roc_obj)
  print(paste("AUC:", round(auc_val, 4)))
  
  roc_df <- data.frame(
    fpr = 1 - roc_obj$specificities,
    tpr = roc_obj$sensitivities
  )
  
  ggplot(roc_df, aes(x = fpr, y = tpr)) +
    geom_line(color = "blue", size = 1.2) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") +
    labs(title = paste("ROC Curve (AUC =", round(auc_val, 4), ")"),
         x = "False Positive Rate",
         y = "True Positive Rate") +
    theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
}
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
data <- read_csv("Bank Customer Churn Prediction.csv") %>%
  janitor::clean_names()

data <- data %>%
  select(-customer_id) %>%
  mutate(churn = factor(churn, levels = c(0, 1), labels = c("no", "yes"))) %>%
  mutate(across(-churn, ~ as.numeric(.))) %>% select(-gender, -country)

set.seed(123)
train_idx <- createDataPartition(data$churn, p = 0.8, list = FALSE)
train_clean <- data[train_idx, ]
test_clean  <- data[-train_idx, ]

train_clean <- train_clean %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), median(., na.rm = TRUE), .)))

cat_vars <- data %>% select(where(is.factor))

cat_long <- cat_vars %>%
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value")
```

# Exploratory Data Analysis (EDA)

Both categorical and numeric variables were explored using graphs. A bar chart was made to 
look at the distribution of the churn variable. Most customers stayed with the bank, and only a 
smaller group left, showing an imbalance in the classes. 
Histograms were used to see how numeric variables like balance, age, and credit score are 
distributed. The balance data was skewed because many customers had very low balances (often 
zero), and only a few had high balances. The age data was more evenly spread out, looking 
similar to a normal distribution. The credit score histogram showed that most customers had 
higher scores, causing a slight skew to the left. 
These graphs made it clear if there were unusual data points (outliers) or if certain variables 
needed adjustments, like scaling, before using them in machine learning models. 

```{r, echo=FALSE, warning=FALSE, message=FALSE}
ggplot(cat_long, aes(x = value)) +
  geom_bar(fill = "steelblue") +
  facet_wrap(~ variable, scales = "free", ncol = 3) +
  theme_minimal() +
  labs(title = "Bar Plots of Categorical Variables", x = "Category", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
num_vars <- data %>% select(where(is.numeric))
```
Fig 1. Distribution of Customer Churn 

This bar chart shows the number of customers who stayed versus those who churned.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
num_long <- num_vars %>%
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value")

ggplot(num_long, aes(x = value)) +
  geom_histogram(bins = 30, fill = "lightgreen", color = "black") +
  facet_wrap(~ variable, scales = "free", ncol = 3) +
  theme_minimal() +
  labs(title = "Histograms of Numeric Variables", x = "Value", y = "Count")
```
Fig2. Histograms of Numeric Features 

This set of histograms displays the distribution of key numeric variables, including balance, age, 
and credit score. 

# Modeling and Resampling 

In this section, four different models were built to predict whether customers would churn. 
Cross-validation (5-fold) was used for each model to make sure the results weren't just by 
chance. 

## Logistic Regression

Logistic regression is a simple method that predicts whether a customer leaves or stays based on 
account features. Cross-validation was used when training this model. 
```{r, echo=FALSE, warning=FALSE, message=FALSE}
ctrl_roc <- trainControl(
  method = "cv",
  number = 5,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final"
)

set.seed(123)
logit_model <- train(
  churn ~ ., 
  data = train_clean,
  method = "glm",
  family = "binomial",
  trControl = ctrl_roc,
  metric = "ROC"
)

print(logit_model)
plot_roc_auc(logit_model, test_clean, "churn")
```

The logistic regression model had an AUC score of 0.7512, showing a reasonable ability to 
predict churn accurately. The ROC curve shows how the model balances true and false 
predictions. 

## Decision Tree

A decision tree splits the data based on certain conditions to classify customers. Cross-validation 
was also used to find the best complexity parameter (cp), which helps prevent overfitting. 

```{r, echo=FALSE, warning=FALSE, message=FALSE}
ctrl_roc <- trainControl(
  method = "cv",
  number = 5,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final"
)

set.seed(123)
tree_model <- train(
  churn ~ .,
  data = train_clean,
  method = "rpart",
  trControl = ctrl_roc,
  metric = "ROC",
  tuneLength = 10
)

print(tree_model)
plot_roc_auc(tree_model, test_clean, "churn")
```

The best decision tree model achieved an AUC of 0.7454, a bit lower than logistic regression, 
but still good enough to be useful.

## KNN

KNN predicts churn by looking at similar customers. Cross-validation helped find the best 
number of neighbors (k). 

```{r, echo=FALSE, warning=FALSE, message=FALSE}
set.seed(123)
knn_model <- train(
  churn ~ .,
  data = train_clean,
  method = "knn",
  trControl = ctrl_roc,
  metric = "ROC",
  tuneLength = 10
)


print(knn_model)
plot_roc_auc(knn_model, test_clean, "churn")
```

With an optimal k = 23, the KNN model only reached an AUC of 0.5543, meaning it didn't do a 
very good job of predicting churn accurately. 

## SVM

SVM tries to find the best line (or boundary) separating customers who churn from those who 
stay. The radial basis function (RBF) kernel was chosen because it's good for data that isn’t 
easily separated by a straight line. 

```{r, echo=FALSE, warning=FALSE, message=FALSE}

set.seed(123)
svm_model <- train(
  churn ~ .,
  data = train_clean,
  method = "svmRadial",
  trControl = ctrl_roc,
  metric = "ROC",
  tuneLength = 5
)

print(svm_model)
plot_roc_auc(svm_model, test_clean, "churn")
```

The SVM had the highest AUC (0.792) of all models tested, making it the best at predicting 
customer churn.


# Model Comparision and Visualization

The performance of the four models was compared using ROC curves, which visually show how 
well each model distinguishes between customers who churned and those who did not. 
Here are the AUC results for each model: 

Logistic Regression: AUC = 0.7512 

Decision Tree: AUC = 0.7454 

K-Nearest Neighbors (KNN): AUC = 0.5543 

Support Vector Machine (SVM): AUC = 0.792 

The SVM model performed the best, showing the highest AUC. This means it was most accurate 
in identifying customers likely to churn. Logistic regression and decision tree models also 
performed reasonably well, while KNN clearly performed the worst. 
The strong performance of SVM might be due to its ability to capture more complex 
relationships between variables, especially when the boundary between churn and non-churn 
customers is not clear-cut. On the other hand, the poor performance of KNN might be because 
the data includes variables with different scales and because KNN tends to struggle with datasets 
having many features. 

# Conclusion

This project built and compared four different machine learning models (Logistic Regression, 
Decision Tree, KNN, and SVM) to predict customer churn at a bank. Among the models, SVM 
had the highest AUC (0.792), making it the most accurate at predicting which customers would 
leave. Logistic regression (0.7512) and decision tree (0.7454) models also did well, while KNN 
(0.5543) performed poorly. 
These results supported the initial hypothesis that features like credit score, balance, customer 
activity, and credit card ownership can effectively predict customer churn. The SVM model 
probably performed best because it handled the complexity and nuances in the data better than 
the other methods. 
In the future, additional variables like gender and geography could be included to see if 
predictions can improve further. Also, combining multiple models into an ensemble approach 
might help achieve even better performance. 